Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-07-28 Multiresolution_framework/ltx/response_to_reviews.tex

        \item Section 5 is a review of known results and standard background material from wavelet theory and includes a simple calculation of the Fourier transform of the particular basis used here, which would be more appropriate for an appendix (cf appendix H of the Freestone et al. paper where the Fourier transform of the Gaussian basis functions is calculated).

		\emph{Response here.}

        \item Section 6 also borrows heavily from Dewar et al, e.g. cf equation (58) of this paper with equation (19) of Dewar et al: the state transition matrix is a perturbation of the corresponding matrix used in that earlier paper. The approach used in both papers seems to be the same with an RTS smoother used in the E-step (this was also used and explained in Freestone et al.) and an M step optimizing a cost function, quadratic in the parameters, cf. equation (62) with  equation (29) of Theorem 2 of Dewar et al. In general, the derivations of section 6 and appendix B follow closely or are very similar to those of Dewar et al.

		\emph{\parham{As the reviewer correctly pointed out there are similarities in the derivations herein and that of Dewar et al. Our main aim here is to adapt techniques from control systems theory and machine learning, modify and apply them to the Amari type neural field models. As the result some of the derivations are naturally similar. For example, the Amari type neural field model in this paper is only different to the IDE model in Dewar et al. due to the membrane time constant and the firing rate, as the result the only changes appear in the transition matrix is the presence of the membrane parameter $\xi=1-T_s/\tau$ and the firing rate slope $\varsigma$. The use of the RTS smoother in the E-step of the EM algorithm for state-space models is an effective, standard and widely used approach. In fact, the aim of the IDE representation in a state-space form is to adapt the EM and consequently the RTS smoother. The RTS algorithm is presented only for completeness and it is different than Unscented RTS (for non-linear state-space models) in Freestone et al. The cost function of a linear state-space model is always quadratic in parameters in the M-step ??? in fact it is one of the advantages of this representation, giving a closed form solution for system's parameters. The novel implementation of the M-step in this work significantly simplifies the computational complexity of the EM based algorithm proposed by Dewar et al. as it is shown in Table.~4 of the manuscript, namely from $O(n_x^4n_{\theta}^2)$ to $O(n_x^2n_{\theta}^2)$. This reduction is very significant and crucial for high dimensional systems and specifically for implementing the multi-resolution framework introduced here which otherwise would be possible.} }

@@

        \item Also, Appendix A is the analogue of Appendix D of Freestone et al. and Appendix C includes results known from the literature (as the authors point out).

\emph{\parham{In order to formulate the state-space representation of the multi-resolution neural field model the inner product of the B-spline functions are required. We believe that it is necessary to provide the interested reader with the required tools to implement the algorithm. Appendix D of Freestone et al. and Appendix A here are only similar in the sense that the inner product of two basis functions are calculated. In Appendix D of Freestone et al. two  n-dimensional Gaussians are considered while here the inner product of two B-spline functions is calculated. As for Appendix C we think that it would be informative for the broad audience of NI to include the derivation of some of the well-known formulations in the EM algorithm.} }

@@

    \end{enumerate}

    \section{Reviewer 2}

\begin{enumerate}
    \item As a minor comment, I would suggest changing the notation for the scaling and wavelet coefficients in Eq. (6), for example into normal lowercase `a' and `c', respectively. The $\beta$ might be confused by B-spline notation.

	\emph{\parham{We have changed the scaling and wavelet coefficients into normal lowercase `a' and `c', respectively.}}

@@

\item For the discussion, it might be useful to add a sentence that the basis structure (i.e., non-redundant, versus a (tight) frame representation) is a requirement for this framework.

\emph{\parham{We have added a sentence to Sect 8.1 (last paragraph) for the basis structure requirement.}}

@@

			\item{Eqn 9: the state variables $x$ seem to arise without distinction from $v$}

			\emph{Response here.}

			\item Between eqn 19 and 20 change
			`write the neural field and connectivity kernel decomposition' To `write the connectivity kernel and neural field decomposition'
			To maintain order with eqns 20 and 21.

			\emph{\parham{We have changed the text to `the connectivity kernel and neural field decomposition' to maintain order with eqns 20 and 21.}}

@@

  		\item Figure 8 is referred to in the text out of order

  			\emph{\parham{The cross-referencing has been corrected.}}

			\item Pg 19: `stationary point i.e. global' should be `stationary point, i.e. a global'

			\emph{\parham{We have changed the text to `stationary point, i.e. a global'.}}

### 2012-07-28 Multiresolution_framework/ltx/manuscript.tex
response to R2

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace} 
% \usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
% \usepackage{hyperref}
% \hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}
% \newcommand{\dean}[1]{\textcolor{green}{#1}}
% \newcommand{\parham}[1]{\textcolor{blue}{#1}}
% \newcommand{\vik}[1]{\textcolor{red}{#1}}

@@

Choosing isotropic scaling and wavelet functions allows switching the order of operations with the inner product and convolution in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. This simplifies the computational complexity of the estimation algorithm significantly. Although this is correct for one dimensional B-spline scaling and wavelet functions, it does not hold at higher dimensions as the resulting B-spline wavelets are not isotropic. One alternative could be  
isotropic polyharmonic B-splines scaling functions and wavelets \citep{VanDeVille2005}.
 \subsection{Modification to higher dimensions}
 The reconstruction of the neural field in higher dimensions requires $2^n-1$ wavelets, where $n$ is the dimension of the space. For example a two-dimensional MRA can be implemented using 2-D scaling and wavelet functions built up using the tensor-product approach \citep{Meyer1992}. In this case three wavelet functions are required to  extract fine features of the field at vertical, horizontal and diagonal orientations. The formulations provided in this paper can be easily extended to two dimensions. However, the computational load of the estimation algorithm will be significant. It is important to investigate methods for choosing basis functions to balance the complexity with the reconstruction capability of the model.
\subsection{Extensions to the framework} 
Clearly there is, and will always be, a discrepancy between the dynamics of the model and the experimental observation. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The MRAIDE approach is not limited to neural fields; the framework can be applied to modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others~\citep{Wikle2002,Xu2005}. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge.

@@

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t^b)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1}^b)^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^{b\top}\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t^b\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}+\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}.
\end{align}
Rearranging   Eq.~\eqref{eq:Xi0derivation0} gives
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}.
\end{equation}
Substituting Eq.~\eqref{eq:Xi0derivation1} into Eq.~\eqref{eq:appXi0}, noting the expectation commutes with the sum yields
\begin{equation}
 \boldsymbol\Xi_0=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right).
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by calculating the covariance matrix, $\mathbf P_t^b$ and following the above steps. 
\newpage
% \bibliographystyle{./elsarticle/model2-names}  
\bibliographystyle{model2-names} 
\bibliography{MRAIDE}
\end{document}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace} 
 \usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
% \usepackage{hyperref}
 % \hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}
 \newcommand{\dean}[1]{\textcolor{green}{#1}}
 \newcommand{\parham}[1]{\textcolor{blue}{#1}}
% \newcommand{\vik}[1]{\textcolor{red}{#1}}

@@

\parham{The basis structure (i.e., non-redundant, versus a (tight) frame representation) is a requirement for this framework.                                              }Choosing isotropic scaling and wavelet functions allows switching the order of operations with the inner product and convolution in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. This simplifies the computational complexity of the estimation algorithm significantly. Although this is correct for one dimensional B-spline scaling and wavelet functions, it does not hold at higher dimensions as the resulting B-spline wavelets are not isotropic. One alternative could be  
isotropic polyharmonic B-splines scaling functions and wavelets \citep{VanDeVille2005}.
 \subsection{Modification to higher dimensions}
 The reconstruction of the neural field in higher dimensions requires $2^n-1$ wavelets, where $n$ is the dimension of the space. For example a two-dimensional MRA can be implemented using 2-D scaling and wavelet functions built up using the tensor-product approach \citep{Meyer1992}. In this case three wavelet functions are required to  extract fine features of the field at vertical, horizontal and diagonal orientations. The formulations provided in this paper can be easily extended to two dimensions. However, the computational load of the estimation algorithm will be significant. It is important to investigate methods for choosing basis functions to balance the complexity with the reconstruction capability of the model. \parham{It is also possible to apply other than dyadic subsampling schemes in multiple dimensions, which leads to less wavelet subbands. For example, in 2D, the quincunx scheme is well-known to have better isotropic properties. Unfortunately, in 3D, it is not possible to have a subsampling operator (acting on a Cartesian sampling lattice) that is performing subsampling at equal rate in each dimension at each iteration \citep{VanDeVille2005a}.}
\subsection{Extensions to the framework} 
Clearly there is, and will always be, a discrepancy between the dynamics of the model and the experimental observation. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The MRAIDE approach is not limited to neural fields; the framework can be applied to modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others~\citep{Wikle2002,Xu2005}. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge.

@@

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t^b)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1}^b)^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^{b\top}\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t^b\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}+\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}.
\end{align}
Rearranging   Eq.~\eqref{eq:Xi0derivation0} gives
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}.
\end{equation}
Substituting Eq.~\eqref{eq:Xi0derivation1} into Eq.~\eqref{eq:appXi0}, noting the expectation commutes with the sum yields
\begin{equation}
 \boldsymbol\Xi_0=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right).
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by calculating the covariance matrix, $\mathbf P_t^b$ and following the above steps. 
\newpage
 \bibliographystyle{./elsarticle/model2-names}  
% \bibliographystyle{model2-names} 
\bibliography{MRAIDE}
\end{document}
